Demes: a specification for interchanging complex demography

### 2021-07-28 paper.tex
edits and filling in bullets

\affil[1]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
\affil[2]{FIXME: Aaron's affiliation}
\affil[3]{Department of Molecular and Cellular Biology, University of Arizona}
\affil[4]{Institute of Evolutionary Biology, The University of Edinburgh}
\affil[5]{Computer Technologies Laboratory, ITMO University}
\affil[6]{Big Data Institute, Li Ka Shing Centre for Health Information and
    Discovery, University of Oxford}
\affil[7]{Ecology and Evolutionary Biology, University of California at Irvine}

@@

Inference methods for demography usually output a set of
text files which describe the inferred model,
and nearly all methods use different file formats.
% Thus, any comparison
% between inferences requires these file formats to be parsed, with the user
% being left with the task of unifying the data models.
Inference results are typically reported in publications
via a combination of visual depiction,
a list of key parameters in tabular form and a discussion within the text.
Unfortunately there is often ambiguity in these descriptions and
implementing the precise model inferred for later simulation
is at best tedious and error prone~\citep{ragsdale2020lessons}
and occasionally impossible because of missing information.
The \stdpopsim\ project is a community-maintained catalog of species
information and demographic models designed to provide standardised
population genetic simulations~\citep{adrion2020community}.
The development process for adding demographic models to
\stdpopsim\ tries to avoid errors in model implementation by requiring two
independent implementations of each model
to be equivalent. While this process is
robust and has uncovered both significant
errors~\citep{ragsdale2020lessons} and ambiguities in
published models, it is labour intensive.
It would be far better if published results
were reported in an unambiguous standardised format that could be provided
directly as input to simulators.

Simulation is a core tool in population genetics, and a
large number of different methods have been developed
over the past three decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a custom input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models be defined programmatically~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.][]{ewing2010msms,chen2009fast,staab2015scrm},
and the POPDemog visualisation method~\citep{zhou2018popdemog}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors highly likely.
Input parameter file formats for simulators allow the model specification
to be less terse and (potentially) also allow for documentation in the
form of comments. However, there is no standard format shared across
simulators, and the details of the demographic models are not separated
from other simulation parameters. The APIs use by individual simulators
to define demographic models are also not appropriate as an interchange
standard because of the dependency on both the programming language involved,
and the specific details of the simulator in question.

Here we present ``Demes'', a data model and file format
specification for complex demographic models developed by the PopSim
Consortium. The Demes data model precisely defines key parameters
of populations and their relationships over time in an extensible way,
and provides a way to explicitly encode all information relevant to
demography while avoiding repetition. This data model is implemented
in the widely used YAML format~\citep{ben2009yaml}, which provides a
good balance between human and machine readability.
The specification precisely
defines the required behaviour of implementations, ensuring that there
is no ambiguity of interpretation, and includes both a reference
implementation and an extensive suite of test examples and their expected
output. The initial software ecosystem includes high-quality
Python and C parser implementations, as well as utilities for verification
and visualisation of Demes models, and has been implemented in
several popular inference and simulation methods
(Table~\ref{tab-software}).
We hope that this
data model and file format will be widely adopted by the community,
such that users can expect to simulate directly from
inferred models with little or no programming effort.

@@

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and the structure formally defined using
JSON Schema~\citep{wright2020json}.
The Demes JSON schema rigorously defines the data model,
specifying the hierarchical structure and the properties that are allowed in different
contexts along with their types and permissable ranges. This schema, and its
accompanying documentation, fully describe the entities in the model and their
relationships, and the required behaviour of implementations. Since the
schema is definitive, we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

The design of Demes is a balance between two partially competing requirements:
that (a) models should be easy for humans to understand and manipulate; and (b)
software processing Demes models should be provided with an unambiguous
representation that is straightforward to process. For efficiency of
understanding and avoidance of model specification error, we require a data
representation without redundancy (i.e., repetition of values). But, for the
simplicity of software working with the Demes model (and the avoidance of
programming error, or divergence in interpretations of the specification) it is
preferable to have an explicit representation, in which all relevant values are
readily available. Thus, we consider three entities: the high-level, human
readable, model specification; the low-level, fully-qualified model
specification; and the parser, which is responsible for transforming the former
into the latter.

@@

In Demes, demographic models consist of one or more interacting populations (or
``demes"; to avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, in the understanding that the
terms are interchangeable for our purposes). Populations are defined as a
collection of exchangeable individuals that exist for a specified period of
time, following a well-defined set of rules regarding population sizes, mating
systems, etc. Ancestor/descendant relationships between populations can be
defined, as well as continuous or instantaneous migration between coexisting
populations.

Population and event times are written as units in the past, so that time zero
corresponds to the final generation or ``now'', and event times in the past are
values greater than zero with larger values for events that occur in the more
distant past. While a natural specification for time units is in generations,
other time units are allowed, such as years.
\jkcomment{Why did we choose time to flow in this direction?}

Population sizes are given as the number of individuals.
\jkcomment{As opposed to genomes. Why did we do this?}
Sizes and mating system details are specified for each population within
epochs. Epochs are contiguous time intervals that define
the existence interval of the population. Each epoch specifies the population size
over that interval, which can be a constant value or function defined by start
and end sizes that must remain positive.

Within a population, individuals are assumed to be exchangeable, but
there are also parameters for nonrandom mating, such as selfing or cloning rates,
which give the probability that offspring are generated from one generation to
the next by self-fertilization or cloning of an individual
\textbf{need to precisely specify once this is determined}.
\jkcomment{More detail here. Also, why only selfing and cloning, why not selection
coefficients and whatnot?}

@@

\begin{figure}
\begin{minipage}{0.48\textwidth}
\begin{tcolorbox}
\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
\end{tcolorbox}
\begin{tcolorbox}
\includegraphics[width=\textwidth]{fig/IM}
\end{tcolorbox}
\end{minipage}\hfill
\begin{minipage}{0.48\textwidth}
\begin{tcolorbox}
\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
\end{tcolorbox}
\end{minipage}\\
\caption{\label{fig-example}
Example isolation-with-migration Demes model. (A) The compact non-redundant
representation expressed using YAML. (B) The same model in the fully-qualified
form. (C) A visual representation of the model using \texttt{demesdraw}.
The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
but are much more human-readable.
}
\end{figure}

@@

Structurally, the high-level Demes specification encourages human understanding
by avoiding redundancy in the description where possible, and
by providing a mechanism for specifying default values that
are inherited hierarchically. Figure~\ref{fig-example} shows an example
Demes model expressed in both the high and low-level forms (both in
terms of YAML syntax).

\jkcomment{Finish up this section with a discussion of how we (a)
use default values defined in JSON schema; (b) define more
subtle implicit values in the specification itself; and (c) use the
defaults sections, all explified by Fig 1.}

\subsection*{Parsers}
The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly decreasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem). We therefore define
a standard software entity as part of the specification (the parser),
which performs this task, and which can be shared by programs that
support demes as an input. By having relatively few (ideally, one
per programming language), high-quality Demes parsers available as
libraries the probability of divergences from the standard is
greatly reduced.

The Demes specification precisely defines the required behaviour of
parsers, and we also provide a reference implementation written
in Python to resolve any potential ambiguities and to provide a
helpful template for other implementations. We also provide an
extensive test suite of examples and the expected outputs. In addition,
we have high-quality parser implementations in the Python and C
languages (published under liberal open source licences),
providing a solid foundation for the software ecosystem.

\subsection*{Unambiguous low-level representation}

\jkcomment{Not sure if there's much point in this as a separate section any
more? This could really be said in the previous section.}

Parsers output the low-level description of Demes models,
which is an explicit representation
of the model where all parameters have been assigned values, and
all data constraints have been checked. The output is formally defined
as JSON, but in practise parsers output an object model that
is suitable for the particular programming language, and
that can be used directly by the implementing program.

@@

\begin{itemize}
    \item Why we've avoided turning this into a model of everything, allowing
    for the specification of genome properties like recombination rates,
    etc.

    \item Why we've versioned the specification, and how we've planned
    for future extensions.

\end{itemize}

@@

Reproducibility is a significant problem throughout the
sciences~\citep{baker20161} and various measures have been
proposed to increase the likelihood of researchers being
able to replicate results in the
literature~\citep{munafo2017manifesto}. The most basic requirement
for reproducibility is that we must be able to state precisely what
the result in question is. The lack of standardisation in how
complex demographic models are communicated today, and the lack of
precision in the published model descriptions means that it is difficult
to replicate analyses, or reproduce those models for later simulation.
Thus, we hope that the Demes standard introduced here will be widely adopted
by inference methods and the results reported in publications
routinely included as supplemental material or uploaded to a data
repository in Demes format.

>>>>

\affil[1]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
\affil[2]{Department of Integrative Biology, University of Wisconsin--Madison}
\affil[3]{Department of Molecular and Cellular Biology, University of Arizona}
\affil[4]{Institute of Evolutionary Biology, The University of Edinburgh}
\affil[5]{Computer Technologies Laboratory, ITMO University}
\affil[6]{Big Data Institute, Li Ka Shing Centre for Health Information and
    Discovery, University of Oxford}
\affil[7]{Ecology and Evolutionary Biology, University of California at Irvine}

@@

Inference methods for demography usually output a set of
text files which describe the inferred model,
and nearly all methods use different file formats.
% Thus, any comparison
% between inferences requires these file formats to be parsed, with the user
% being left with the task of unifying the data models.
Inference results are typically reported in publications
via a combination of visual depiction,
a list of key parameters in tabular form and a discussion within the text.
Unfortunately there is often ambiguity in these descriptions, and
implementing the precise model inferred for later simulation
is at best tedious and error prone~\citep{ragsdale2020lessons}
and occasionally impossible because of missing information.
The \stdpopsim\ project is a community-maintained catalog of species
information and demographic models designed to provide standardised
population genetic simulations~\citep{adrion2020community}.
The development process for adding demographic models to
\stdpopsim\ tries to avoid errors in model implementation by requiring two
independent implementations of each model
to be equivalent. While this process is
robust and has uncovered both significant
errors~\citep{ragsdale2020lessons} and ambiguities in
published models, it is labour intensive.
It would be far better if published results
were reported in an unambiguous standardised format that could be provided
directly as input to simulators.

Simulation is a core tool in population genetics, and a
large number of different methods have been developed
over the past three decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer}.
Simulations are based on highly idealised population models
and one of the key uses of inferred demographic histories
is to make simulations more realistic.
Simulation methods take three broad approches to specifying
the demographic model to simulate,
using either a command line
interface~\citep[e.g.][]{hudson2002generating,kern2016discoal},
a custom input file
format~\citep[e.g.][]{guillaume2006nemo,excoffier2011fastsimcoal,shlyakhter2014cosi2},
or an Application Programming Interface (API) to allow
models be defined programmatically~\citep[e.g.][]{
thornton2014cpp,kelleher2016efficient,becheler2019quetzal,haller2019slim}.
Command line interfaces are a concise way of expressing
demographic models, and the syntax defined by \ms~\citep{hudson2002generating}
is used by a number of different
simulators~\citep[e.g.][]{ewing2010msms,chen2009fast,staab2015scrm},
and the POPDemog visualisation method~\citep{zhou2018popdemog}.
However, this conciseness means that models of even intermediate complexity
are difficult for humans to understand, making errors highly likely.
Input parameter file formats for simulators allow the model specification
to be less terse and (potentially) also allow for documentation in the
form of comments. However, there is no standard format shared across
simulators, and the details of the demographic models are not separated
from other simulation parameters. The APIs used by individual simulators
to define demographic models are also not appropriate as an interchange
standard because of the dependency on both the programming language involved,
and the specific details of the simulator in question.

Here we present ``Demes'', a data model and file format
specification for complex demographic models developed by the PopSim
Consortium. \aprcomment{do we need ``developed by the PopSim Consortium''?}
The Demes data model precisely defines key parameters
of populations and their relationships over time in an extensible way,
and provides a way to explicitly encode all information relevant to
demography while avoiding repetition. This data model is implemented
in the widely used YAML format~\citep{ben2009yaml}, which provides a
good balance between human and machine readability.
The specification precisely
defines the required behaviour of implementations, ensuring that there
is no ambiguity of interpretation, and includes both a reference
implementation and an extensive suite of test examples and their expected
output. The initial software ecosystem includes high-quality
Python and C parser implementations, as well as utilities for verification
and visualisation of Demes models, and has been implemented in
several popular inference and simulation methods
(Table~\ref{tab-software}).
We hope that this
data model and file format will be widely adopted by the community,
such that users can expect to simulate directly from
inferred models with little or no programming effort.

@@

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and the structure formally defined using
JSON Schema~\citep{wright2020json}.
The Demes JSON schema rigorously defines the data model,
specifying the hierarchical structure and the properties that are allowed in different
contexts along with their types and permissable ranges. This schema, and its
accompanying documentation, fully describe the entities in the model and their
relationships and the required behaviour of implementations. Since the
schema is definitive, we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

The design of Demes is a balance between two partially competing requirements:
that (a) models should be easy for humans to understand and manipulate; and (b)
software processing Demes models should be provided with an unambiguous
representation that is straightforward to process. For efficiency of
understanding and avoidance of model specification error, we require a data
representation without redundancy (i.e., repetition of values). But, for the
simplicity of software working with the Demes model (and the avoidance of
programming error, or divergence in interpretations of the specification) it is
preferable to have an explicit representation, in which all relevant values are
readily available. Thus, we consider three entities: the high-level, human
readable model specification; the low-level, fully-qualified model
specification; and the parser, which is responsible for transforming the former
into the latter.

@@

In Demes, demographic models consist of one or more interacting populations (or
``demes''; to avoid confusion with the name of the specification itself we will
use the term ``population'' in this discussion, in the understanding that the
terms are interchangeable for our purposes). Populations are defined as a
collection of exchangeable individuals that exist for a specified period of
time, following a well-defined set of rules regarding population sizes, mating
systems, and so on. Ancestor/descendant relationships between populations can
be defined, as well as continuous or instantaneous migration between coexisting
populations.

Population and event times are written as units in the past, so that time zero
corresponds to the final generation or ``now'', and event times in the past are
values greater than zero with larger values for events that occur in the more
distant past. By having time units increase into the past, we avoid the need to
choose an arbitrary point in history as ``time zero'', so that instead zero
refers to ``now'' or the final generation of a simulation. A natural
specification for time units is in generations, although other time units are
permitted, such as years, accompanied by the generation time.

Population sizes are given as the number of individuals.
\jkcomment{As opposed to genomes. Why did we do this?}
\aprcomment{With this, we might want to be able to specify ploidy in as
a graph-level option, which defaults to 2. That would make the reasons
behind this decision obvious then, I think.}
Sizes and mating system details are specified for each population within
epochs. Epochs are contiguous time intervals that define
the existence interval of the population. Each epoch specifies the population size
over that interval, which can be a constant value or function defined by start
and end sizes that must remain positive.

Within a population, individuals are assumed to be exchangeable, but there are
also parameters for nonrandom mating, such as selfing or cloning rates, which
give the probability that offspring are generated from one generation to the
next by self-fertilization or cloning of an individual. Cloning and selfing
rates can take any value between zero and one, and if both are specified then
the selfing rate is the probability an offspring is generated through
self-fertilization given it was produced sexually (i.e. with probability one
minus the cloning rate). We omit any details regarding selection, as those would
fall outside the scope of demographic model specification, but...
\aprcomment{If selection is placed on top of the demographic model, this implies
that we are living in a soft selection world. Though sizes could just as easily
be interpreted as carrying capacities in a hard selection model.}

\jkcomment{More detail here. Also, why only selfing and cloning, why not selection
coefficients and whatnot?}

@@

\begin{figure}
\begin{minipage}{0.44\textwidth}
\begin{tcolorbox}
\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
\end{tcolorbox}
\begin{tcolorbox}
\includegraphics[width=\textwidth]{fig/IM}
\end{tcolorbox}
\end{minipage}\hfill
\begin{minipage}{0.54\textwidth}
\begin{tcolorbox}
\inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
\end{tcolorbox}
\end{minipage}\\
\caption{\label{fig-example}
Example isolation-with-migration Demes model. (A) The compact non-redundant
representation expressed using YAML. (B) The same model in the fully-qualified
form. (C) A visual representation of the model using \texttt{demesdraw}.
The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
but are much more human-readable.
}
\end{figure}

@@

Structurally, the high-level Demes specification encourages human understanding
by avoiding redundancy in the description where possible and by providing a
mechanism for specifying default values that are inherited hierarchically.
Figure~\ref{fig-example} shows an example Demes model expressed in both the
high and low-level forms (both in terms of YAML syntax). For values that repeat
across fields, default may be used to implicity assign a default value to
fields of the given type.  A default is superceded by an explicitly provided
value if given. Other values are inherited naturally following the progression
of time. For example, if an epoch \texttt{start\_size} is not provided, it is
assumed to be equal to the \texttt{end\_size} of the previous epoch. This also
means that the first epoch of each population must specify the initial size.

\aprcomment{not sure if this is sufficient here}

\subsection*{Parsers}
The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly decreasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem). We therefore define
a standard software entity as part of the specification (the parser),
which performs this task, and which can be shared by programs that
support demes as an input. By having relatively few high-quality Demes
parsers available as libraries (ideally, one per programming language),
the probability of divergences from the standard is greatly reduced.

The Demes specification precisely defines the required behaviour of parsers,
which translate the high-level specifications into unambiguous low-level
representations where all parameters have been assigned values and data
constraints have been checked. The output is formally defined as JSON, but in
practise parsers output an object model that is suitable for the particular
programming language and that can be used directly by the implementing program.
We provide a reference implementation written in Python to resolve any
potential ambiguities and to provide a helpful template for other
implementations, as well as an extensive test suite of examples and the
expected outputs. In addition, we have high-quality parser implementations in
the Python and C languages (published under liberal open source licences),
providing a solid foundation for the software ecosystem.

\aprcomment{wrapped below section about parsers to here}

@@

Demes is meant to interface with a large number of simulation and inference
methods, so we have restricted to the specification to only describe
demographic features at the population level. As the underlying discrete
demographic structure is the common thread connecting many widely used
population genetic software, many features of underlying genome biology are
omitted.  Most notably, mutation and recombination rates are left unspecified
(though they can be encoded as metadata in the data structure). Genome ploidy
is left ambiguous, and selection and dominance models are absent. Using a Demes
model, which specifies deterministic population sizes, as the demographic input
for simulation using software that allows for selection implies that selection
is \emph{soft}, that is, genome fitnesses are rescaled relative to the mean
population fitness \citep{christiansen1975hard}. It is possible to instead
interpret population sizes as carrying capacities, allowing for a model of
\emph{hard} selection. However, this should not be understood as the default
interpretation of population sizes, which conventionally specifies a
deterministic number of individuals.  Finally, as Demes assumes the
exchangeability of individuals within populations, Demes does not provide a way
to define geographic spatial structure within populations.

Demes version 1.0 is intended to provide a stable and minimal specification
that is compatible with a wide range of simulation softwares. Beginning with
version 1.0, a Demes-specified model is expected to provide the version of
Demes for which it is valid, and if unprovided, it should be a valid model
under the default version, 1.0. Future extensions of the Demes specification
will be released with versions following semantic versioning guidelines [cite
or url?], and downstream software that incorporate Demes are expected to
indicate and check Demes version compatibility.

@@

Reproducibility is a significant problem throughout the
sciences~\citep{baker20161} and various measures have been
proposed to increase the likelihood of researchers being
able to replicate results in the
literature~\citep{munafo2017manifesto}. The most basic requirement
for reproducibility is that we must be able to state precisely what
the result in question is. The lack of standardisation in how
complex demographic models are communicated today, and the lack of
precision in the published model descriptions means that it is difficult
to replicate analyses, or reproduce those models for later simulation.
Thus, we hope that the Demes standard introduced here will be widely adopted
by simulation and inference methods and be used for reporting results in
publications, either as supplemental material or uploaded to a data
repository in Demes format.
